# Using R to analyse occupation as a function of geographical variables using logistic regression.

Ariane Burke 

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "06_Burke")
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

## Introduction

This demonstration is based on research published in:

Burke, Ariane, et al. 2014	Exploring the impact of climate variability during the Last Glacial Maximum on the pattern of human occupation of Iberia. Journal of Human Evolution 73:35-46. http://dx.doi.org/10.1016/j.jhevol.2014.06.003

```{r}
## ----- Install  packages 
require(car)  ## for vif (variance inflation test)
# BM # we can install coefplot2 from:
# devtools::install_github("palday/coefplot2", subdir = "pkg") # or 
# install.packages("coefplot2",repos="http://www.math.mcmaster.ca/bolker/R",     type="source")
require(AICcmodavg) ## for AIC model selection
require(MASS) ## for stepAIC
library(coefplot2) ## for coefficient plot.
```


```{r fig.show='hold'}
## Set file name
fileName <- "Test_Iberia.csv"

## ----- Read in data
dat <- read.csv(fileName, header=T) 

## ----- View/check data
head(dat)

library(ggplot2)
library(viridis)
map_elev <- 
ggplot(dat, 
       aes(X, 
           Y,
           colour = elev)) +
  geom_point(size = 2,
             shape = 15) +
  scale_color_viridis() +
  theme_minimal() + 
  coord_equal() +
  ggtitle("Study area with elevation")

map_precip  <- 
ggplot(dat, 
       aes(X, 
           Y,
           colour = p_min_spr)) +
  geom_point(size = 2,
             shape = 15) +
    scale_color_viridis(option = "C") +
  theme_minimal() + 
  coord_equal() +
  ggtitle("Study area with precipitation")

library(gridExtra)
grid.arrange(map_elev, map_precip)
```


```{r}
###  ----- Standardization of predictors
dat1 <- as.data.frame(cbind(dat[,c(1:4)], (scale(dat[,-c(1:4)]))))
#dat2<- dat2[,] ## removal of extra non informative variables

# Verification (make sure the vars are numerical)
str(dat1)
```

Using all of the data including the zeros, which cannot be described as necessarily "true" zeroes, we introduce an artificial level of certainty regarding absences. This can bias results (King and Zeng, 2000; Dixon et al.,2005), (also see http://www2.unil.ch/biomapper/Download/Chefaoui-EcoMod-2008.pdf)

To reduce this uncertainy, we sample the zeros which we now call pseudo absences. Good practice is 10x the number of presences (=1000). We can bootstrap this if we want, but this 1000 samples should be sufficient. We will nonetheless need to report some sensitivity to the number of absences used. 

```{r}
## ----- Tease apart presences and absences, then select a subset of absences - recombine presences (all) and absences (sample)
numAbsences <- 350 ## 10x presences

Presences <- subset(dat1, dat1$LGM == 1)
Absences <- subset(dat1, dat1$LGM == 0)[sample(c(1:dim(dat1)[1]), numAbsences),]

## ----- Final data on which analyses will be run
dat2 <- rbind(Presences, Absences)

	## ----- data table for forward selection (stepAIC)
dat3 <- dat2[,c(2, 5, 6, 7, 8, 9, 10)]
	## ----- View/check data
head(dat3)
```

```{r}
## ----- MODELLING

## ----- Define formulae
form0 <- formula(LGM ~ 1) #intercept only model
form1 <- formula(LGM ~ Y + X)
form2 <- formula(LGM ~ elev + slope)
form3 <- formula(LGM ~ t_min_y)
form4 <- formula(LGM ~ elev + slope + p_min_spr)
form5 <- formula(LGM ~ p_min_spr + t_min_y)
form6 <- formula(LGM ~ t_avg_y + p_avg_y)
form7 <- formula(LGM ~ elev + slope + t_avg_y + p_avg_y)
form8 <- formula(LGM~ .) ## all variables for step-wise procedure
## all variables for step-wise procedure
```


```{r}
## ----- Build models
mod.0  <-  glm(form0, family = binomial, data = dat2)
mod.1  <-  glm(form1, family = binomial, data = dat2)
mod.2  <-  glm(form2, family = binomial, data = dat2)
mod.3  <-  glm(form3, family = binomial, data = dat2)
mod.4  <-  glm(form4, family = binomial, data = dat2)
mod.5  <-  glm(form5, family = binomial, data = dat2)
mod.6  <-  glm(form6, family = binomial, data = dat2)
mod.7  <-  glm(form7, family = binomial, data = dat2)
mod.8 <-   stepAIC(glm(form8, family = binomial, data = dat3))
```




```{r}
## ----- Summarize AIC results, including weightings. Using modaicavg package.
mods <-
  list(mod.0,
       mod.1, 
       mod.2, 
       mod.3, 
       mod.4, 
       mod.5, 
       mod.6, 
       mod.7, 
       mod.8)

modnames <-
  c("mod.0",
  "mod.1",
  "mod.2",
  "mod.3",
  "mod.4",
  "mod.5",
  "mod.6",
  "mod.7",
  "mod.8")

aictab(mods, modnames, second.ord = T)

summary(mod.8)
```

```{r}
## ---- Coefficient Plot for model 


coefplot2(mod.8,
          main = "Model 8", 
          col = 'blue', 
          cex.pts = 1.3,
          intercept = FALSE)

## ---- list coefficients
rownames(summary(mod.8)$coefficients)
```

```{r}
## ---- Odds ratios and 95% CI
ORs<- exp(cbind(OR = coef(mod.8), confint(mod.8)))[-1,] ## Intercept OR shouldn't be interpreted.
ORs

## ----- Assess variance inflation (>5 is not good) 
vif(mod.8)

```


```{r}
sessionInfo()
```
