---
title: "Landscape-based Hypothesis Testing in *R*"
author: "Kyle Bocinsky"
date: "3/26/2017"
output:
  html_document:
    code_folding: show
  html_notebook: default
# csl: journal-of-archaeological-science-reports.csl
bibliography: references.bib
---

## Introduction
Many region-scale analyses in archaeology begin with a simple question: How do site locations relate to landscape attributes, such as elevation, soil type, or distance to water or other resources. Such a question is the foundation of basic geospatial exploratory data analysis, and answering it for a set of landscape attributes is the first step towards doing more interesting things, from interpreting settlement patterns in the past [@Kowalewski2008], to the construction of sophisticated predictive models of site location [@Graves2012;@Kohler1986;@Maschner1995], to guiding settlement decision frameworks in agent-based simulations [e.g., @Axtell2002;@Griffin2007;@Kohler2012]. **In this tutorial, we will learn how to use *R* to load, view, and explore site location data, and perform a very basic statistical settlement pattern analysis relating site location to elevation.**

Of course, archaeological site locations are often sensitive information, and it wouldn't be prudent to provide them in tutorial like this. So instead of using cultural site locations, we'll use a point dataset for which we can make a reasonable hypothesis concerning landscape attributes: [cell tower locations from the US Federal Communications Commission](http://wireless.fcc.gov/uls/index.htm?job=transaction&page=weekly). The cell tower data are somewhat difficult to work with, so I've distilled a snapshot of the database (accessed on February 14, 2017), and posted it online for easy download. We'll go through the process of downloading them later in this tutorial. The hypothesis we'll be testing is that cell towers are positioned in unusually high places on the landscape. This is similar to hypotheses we might make about archaeological site locations, perhaps having to do with defensibility [e.g., @Bocinsky2014;@Martindale2009;@Sakaguchi2010] or intervisibility and signaling [e.g., @Johnson2003;@VanDyke2016].

This tutorial is an *R Markdown* HTML document, meaning that all of the code to perform the calculations presented here **was run when this web page was built**---the paper was *compiled*. "Executable papers" such as this one are fantastic for presenting reproducible research in such a way that **data, analysis, and interpretation** are each given equal importance. Feel free to use this analysis as a template for your own work. All data and code for performing this analysis are available on Github at [https://github.com/bocinsky/r_tutorials](https://github.com/bocinsky/r_tutorials).

## Learning goals
In this short tutorial, you will learn how to:
  
  - Download files from the internet in *R*
  - Read ESRI shapefiles and other spatial objects into *R* using the [`sf`](https://github.com/edzer/sfr) and `raster` packages
  - Promote tabular data to spatial objects
  - Crop spatial objects by overlaying them atop one another
  - Generate interactive web-maps using the [`leaflet`](https://rstudio.github.io/leaflet/) package
  - Download federated datasets (including elevation data) using the [`FedData`](http://www.bocinsky.io/FedData/) package
  - Extract data from a raster for specific points
  - Calculate and graph Monte Carlo sub-sampled kernel density estimates for random locations
  - Calculate Monte Carlo sub-sampled Mann-Whitney U test statistics (a non-parametric equivalent to the Student's t-test)
  
We will also gloss over several other fun topics in *R* data analysis, so [download the source](https://github.com/bocinsky/r_tutorials) to learn more!

## Loading necessary packages
Almost all *R* scripts require that you load packages that enable the functions you use in your script. We load our requisite packages here using the `install_cran()` function from the [*devtools*](https://github.com/hadley/devtools) library. `install_cran()` is nice because it will checck whether we have the most recent version of the packages, and only install the ones that are necessary. We also use the `walk()` function from the [*purrr*](http://purrr.tidyverse.org/) package to quickly traverse a vector, but that's a topic for another tutorial.

```{r setup, results='hide', message=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packages <- c("magrittr", # For piping
              "foreach", "purrr", "tibble", "dplyr", "tidyr", "broom", # For tidy data analysis
              "ggplot2","plotly", # For fancy graphs
              "sf", "sp", "raster", # For spatial analysis
              "leaflet", "htmltools", # For fancy maps
              "FedData")

# install.packages("devtools")
# devtools::install_cran(packages, repos = "https://cran.rstudio.com/") # For downloading federated geospatial data
purrr::walk(packages,
            library,
            character.only = TRUE) # Load all packages with no output

# Create an output folder in our current working directory
dir.create("OUTPUT",
           showWarnings = FALSE,
           recursive = FALSE)
```

## Defining the study area
All landscape-scale analyses start with the definition of a study area. Since the cell tower dataset with which we'll be working covers the whole USA, we could really set our study area to be anywhere in the US. Here, we will download an 1:500,500 scale ESRI shapefile of counties in the United States available from the US Census, and pick a county to serve as our study area. I'm going to pick Whitman county, Washington, because that's where I live; feel free to choose your own county!

Files may be downloaded in *R* using many different functions, but perhaps the most straightforward is the `download.file()` function, which requires that you specify a `url` to the file you wish to download, and a `destfile` where the downloaded file should end up. As the counties shapefile is in a zip archive, we will also use the `unzip()` function, which requires you define an extraction directory (`exdir`).

```{r message=FALSE, results='hide'}
# Download the 1:500000 scale counties shapefile from the US Census
if(!file.exists("OUTPUT/cb_2015_us_county_500k.zip"))
  download.file("http://www2.census.gov/geo/tiger/GENZ2015/shp/cb_2015_us_county_500k.zip",
                destfile = "OUTPUT/cb_2015_us_county_500k.zip")

# Unzip the file
unzip("OUTPUT/cb_2015_us_county_500k.zip",
      exdir = "OUTPUT/counties")

```

Navigate to the `exdir` you specified and check to make sure the shapefile is there.

Now it's time to load the shapefile into *R*. We'll be using the `st_read()` function from the [*sf*](https://github.com/edzer/sfr) library, which reads a shapefile (and many other file formats) and stores it in memory as a "simple feature" object of the *sf* library. To read a shapefile, the `st_read()` function requires only the path to the shapefile with the ".shp" file extension. Other spatial file formats have different requirements for `st_read()`, so read the documentation (`help(st_read)`) for more information.

```{r}
# Load the shapefile
census_counties <- sf::st_read("OUTPUT/counties/cb_2015_us_county_500k.shp")

# Inspect the spatial object
head(census_counties)

```

When we inspect the `census_counties` object, we see that it is a simple feature collection with `r nrow(census_counties)` features. Simple feature collection objects are simply data frames with an extra column for spatial objects, and metadata including projection information.

Now it's time to extract just the county we want to define our study area. Because a Simple feature collection object *extends* the `data.frame` class, we can perform selection just as we would with a `data.frame`. We do that here using the `filter()` function from the [*dplyr*](https://github.com/hadley/dplyr) package:

```{r}
# Select Whitman county
my_county <- dplyr::filter(census_counties, NAME == "Whitman")

# Inspect the spatial object
my_county
```

As you can see, the spatial object now has only one feature, and it is `r my_county$NAME` county! We'll map it in a minute, but first let's do two more things to make our lives easier down the road. We'll be mapping using the [*leaflet*](https://rstudio.github.io/leaflet/) package, which makes pretty, interactive web maps. For *leaflet*, we need the spatial data to be in geographic coordinates (longitude/latitude) using the WGS84 ellipsoid. Here, we'll transform our county to that projection system using the `st_transform()` function.

This code chunk also uses something new: the **pipe** operator `%>%` from the [*magrittr*](https://cran.r-project.org/web/packages/magrittr/vignettes/magrittr.html) package. The pipe operator enables you to "pipe" a value forward into an expression or function call---whatever is on the left hand side of the pipe becomes the first argument to the function on the right hand side. So, for example, to find the mean of the numeric vector `c(1,2,3,5)` by typing `mean(c(1,2,3,5))`, we could instead use the pipe: `c(1,2,3,5) %>% mean()`. Try running both versions; you should get `r c(1,2,3,5) %>% mean()` for each. The pipe isn't much use for such a simple example, but becomes *really* helpful for code readability when chaining together many different functions. The compound assignment operator `%<>%` pipes an object forward into a function or call expression and then updates the left hand side object with the resulting value, and is equivalent to `x <- x %>% fun()`

```{r}

# Transform to geographic coordinates
my_county %<>%
  sf::st_transform("+proj=longlat +datum=WGS84")

```

## Reading "site" locations from a table and cropping to a study area
Alright, now that we've got our study area defined, we can load our "site" data (the cell towers). We can use the `read_csv()` function from the [*readr*](http://readr.tidyverse.org/) library to read the cell tower locations from a local file, which we download using `download.file()`. We'll read them and then print them.

```{r message=FALSE}
# Download cell tower location data
if(!file.exists("OUTPUT/cell_towers.csv"))
  download.file("https://raw.githubusercontent.com/bocinsky/r_tutorials/master/data/cell_towers.csv",
                destfile = "OUTPUT/cell_towers.csv")

# Load the data
cell_towers <- readr::read_csv("OUTPUT/cell_towers.csv")

cell_towers
```

As you can see, the cell tower data includes basic identification information as well as geographic coordinates in longitude and latitude. We can use the coordinate data to *promote* the data frame to a spatial object using the `st_as_sf()` function. Finally, we can write a simple function to select only the cell towers in our study area using the `filter()` function again, along with the `st_within()` function from the *sf* package.

```{r message=FALSE, warning=FALSE}
# Promote to a simple feature collection
cell_towers %<>%
  sf::st_as_sf(coords = c("Longitude","Latitude"),
               crs = "+proj=longlat +datum=WGS84")

# Select cell towers in our study area
cell_towers %<>%
  dplyr::filter({
    sf::st_within(., my_county, sparse = F) %>% # This function returns a matrix
                  as.vector() # coerce to a vector
    })

cell_towers
```

Now we see that there are `r nrow(cell_towers)` cell towers in our study area.

## Visualizing site locations

There are *many* different ways to visualize spatial data in *R*, but perhaps the most useful is using the [*leaflet*](https://rstudio.github.io/leaflet/) package, which allows you to make interactive HTML maps that will impress your friends, are intuitive to a 4-year-old, and will thoroughly confuse your advisor. I'm not going to go through all of the syntax here, but in general *leaflet* allows you to layer spatial objects over open-source map tiles to create pretty, interactive maps. Here, we'll initialize a map, add several external base layer tiles, and then overlay our county extent and cell tower objects. *leaflet* is provided by the good folks at [RStudio](https://www.rstudio.com/), and is well documented [here](https://rstudio.github.io/leaflet/). Zoom in on the satellite view, and you can see the cell towers!

```{r message=FALSE}
# Create a quick plot of the locations
leaflet(width = "100%") %>% # This line initializes the leaflet map, and sets the width of the map at 100% of the window
  addProviderTiles("OpenTopoMap", group = "Topo") %>% # This line adds the topographic map from Garmin
  addProviderTiles("OpenStreetMap.BlackAndWhite", group = "OpenStreetMap") %>% # This line adds the OpenStreetMap tiles
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>% # This line adds orthoimagery from ESRI
  addProviderTiles("Stamen.TonerLines", # This line and the next adds roads and labels to the orthoimagery layer
                   group = "Satellite") %>%
  addProviderTiles("Stamen.TonerLabels",
                   group = "Satellite") %>%
  addPolygons(data = my_county, # This line adds the Whitman county polygon
              label = "My County",
              fill = FALSE,
              color = "red") %>%
  addMarkers(data = cell_towers,
             popup = ~htmlEscape(`Entity Name`)) %>% # This line adds cell tower locations
  addLayersControl( # This line adds a controller for the background layers
    baseGroups = c("Topo", "OpenStreetMap", "Satellite"),
    options = layersControlOptions(collapsed = FALSE),
    position = "topleft")
```

It's obvious from this map that the cell towers aren't merely situated to be in high places---they also tend to cluster along roads and in densely populated areas.

## Downloading landscape data with *FedData*
[*FedData*](http://www.bocinsky.io/FedData/) is an *R* package that is designed to take much of the pain out of downloading and preparing data from federated geospatial databases. For an area of interest (AOI) that you specify, each function in *FedData* will **download** the requisite raw data, **crop** the data to your AOI, and **mosaic** the data, including merging any tabular data. Currently, *FedData* has functions to download and prepare these datasets:

  * The [**National Elevation Dataset (NED)**](http://ned.usgs.gov) digital elevation models (1 and 1/3 arc-second; USGS)
  * The [**National Hydrography Dataset (NHD)**](http://nhd.usgs.gov) (USGS)
  * The [**Soil Survey Geographic (SSURGO) database**](http://websoilsurvey.sc.egov.usda.gov/) from the National Cooperative Soil Survey (NCSS), which is led by the Natural Resources Conservation Service (NRCS) under the USDA,
  * The [**Global Historical Climatology Network (GHCN)**](http://www.ncdc.noaa.gov/data-access/land-based-station-data/land-based-datasets/global-historical-climatology-network-ghcn) daily weather data, coordinated by the National Oceanic and Atmospheric Administration (NOAA),
  * The [**Daymet**](https://daymet.ornl.gov/) gridded estimates of daily weather parameters for North America, version 3, available from the Oak Ridge National Laboratory's Distributed Active Archive Center (DAAC), and
  * The [**International Tree Ring Data Bank (ITRDB)**](http://www.ncdc.noaa.gov/data-access/paleoclimatology-data/datasets/tree-ring), coordinated by NOAA.
  
In this analysis, we'll be downloading the 1 arc-second elevation data from the NED under our study area. The *FedData* functions each require four basic parameters:

  * A `template` defining your AOI, supplied as a [`spatial*`](https://www.rdocumentation.org/packages/sp/topics/sp), [`raster*`](https://www.rdocumentation.org/packages/raster/topics/raster), or spatial [`extent`](https://www.rdocumentation.org/packages/raster/topics/extent) object
  * A character string (`label`) identifying your AOI, used for file names
  * A character string (`raw.dir`) defining where you want the raw data to be stored; this will be created if necessary
  * A character string (`extraction.dir`) defining where you want the extracted data for your AOI to be stored; this will also be created if necessary
  
Here, we'll download the 1 arc-second NED with the `get_ned()` function from *FedData*, using the `my_county` object that we created above as out `template`, and local relative paths for our `raw.dir` and `extraction.dir`. We'll download and prepare the NED, and then plot it using the basic `plot()` function.
  
```{r}
# Download the 1 arc-second NED elevation model for our study area
my_county_NED <- FedData::get_ned(template = my_county %>%
                                    as("Spatial"), # FedData functions currently expect Spatial* objects, so coerce to a Spatial* object
                                  label = "my_county",
                                  raw.dir = "OUTPUT/RAW/NED",
                                  extraction.dir = "OUTPUT/EXTRACTIONS/NED")

# Mask to the county
my_county_NED[is.na(over(my_county_NED %>%
                           as("SpatialPixels"),
                         my_county %$%
                           geometry %>%
                           sf::st_transform(projection(my_county_NED)) %>%
                           as("Spatial")))] <- NA

# Print the my_county_NED object
my_county_NED

# Plot the my_county_NED object
my_county_NED %>%
  plot()

# Plot the my_county polygon over the elevation raster
my_county %$% 
  geometry %>%
  plot(add = T)

```

The NED elevation data was downloaded for our study area, **cropped** to the rectangular extent of the county, and then **masked** to the county.

## Are sites situated based on elevation?
Alright, now that all of our data are in place, we can get to the meat of our analysis: are cell towers in especially high places in Whitman county? Here, we treat Whitman county as the [decision space](https://en.wikipedia.org/wiki/Multiple-criteria_decision_analysis) within which some past cellular engineers decided to build towers---our null hypothesis is that there is no relationship between elevation and cell tower location. Put another way, our task is to rule out whether cell towers were likely to have been placed randomly across the landscape. We'll first **extract** the elevations for the cell towers, then calculate a **probability density estimate** for the cell towers, estimate a probability density curve for the landscape using **Monte Carlo resampling**, and finally compare the two distributions graphically and numerically using a **Mann-Whitney U test**.

### Extract site elevations
Extracting the cell tower elevations is straightforward using the `extract()` function from the *raster* package.
```{r}
# Extract cell tower elevations from the study area NED values
cell_towers$`Elevation (m)` <- my_county_NED %>%
  raster::extract(cell_towers %>%
                    as("Spatial"))

cell_towers
```

### Calculate kernel density curves for sites and landscape
We can calculate kernel density curves using the `density()` function available in all *R* installations. This code block gets a little complicated. The first section is strightforward: we estimate the probability density for all elevations between 150 and 1250 masl (the domain of the county elevation). The second section is a bit more complicated: we estimate probability densities for 99 random samples from the elevation data. (You would probably want to draw more resamples than this in a real analysis). Each sample is of the same number of sites as there are cell towers. This is called Monte Carlo resampling. The code section performs the sampling, then calculates a 95% confidence interval for the sampled data using [quantiles](https://en.wikipedia.org/wiki/Quantile). We will use the [*foreach*](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.pdf) package (and its `%do%` function) to repeat and output the resampling.
```{r}

# Calculate the cell tower densities
# -------------------------
cell_towers_densities <- cell_towers %$%
  `Elevation (m)` %>%
  density(from = 150,
            to = 1250,
            n = 1101) %>% 
    tidy() %>%
    tibble::as_tibble() %>%
  dplyr::mutate(y = y * 1101) %>%
  dplyr::rename(Elevation = x,
                Frequency = y)


# Calculate possible densities across the study area using resampling
# -------------------------
# Load the NED elevations into memory for fast resampling
my_county_NED_values <- my_county_NED %>%
  values() %>%
  na.omit() # Drop all masked (NA) locations

# Draw 99 random samples, and calculate their densities
my_county_NED_densities <- foreach::foreach(n = 1:99, .combine = rbind) %do% {
  my_county_NED_values %>%
    sample(nrow(cell_towers),
           replace = FALSE) %>%
    density(from = 150,
            to = 1250,
            n = 1101) %>% 
    broom::tidy() %>%
    tibble::as_tibble() %>%
    dplyr::mutate(y = y * 1101)
} %>%
  dplyr::group_by(x) %>%
  purrr::by_slice(function(x){
    quantile(x$y, probs = c(0.025, 0.5, 0.975)) %>%
      t() %>%
      broom::tidy()
  }, .collate = "rows") %>%
  magrittr::set_names(c("Elevation", "Lower CI", "Frequency", "Upper CI"))

```

### Plot the kernel density curves
We'll perform a statistical test on the cell tower and resampled elevation data in a minute, but first it is just as helpful to view a graph of the two data sets. Like all things, *R* has many different ways of graphing data, but the [*ggplot2*](http://ggplot2.org/) package from [Hadley Wickam](http://hadley.nz/) is fast becoming the framework-*du jour* for graphics in *R*. The [*plotly*](https://plot.ly/r/) package can be used to effortly convert `ggplot2` graphs into interactive HTML graphics. *ggplot2* uses a pipe-like system for building graphs, where graphical elements are appended to one-another using the `+` operator. **Hover over the plot to explore it interactively.**
```{r}
# Plot both distributions using ggplot2, then create an interactive html widget using plotly.
g <- ggplot() +
  geom_line(data = my_county_NED_densities,
            mapping = aes(x = Elevation,
                          y = Frequency)) +
  geom_ribbon(data = my_county_NED_densities,
              mapping = aes(x = Elevation,
                            ymin = `Lower CI`,
                            ymax = `Upper CI`),
              alpha = 0.3) +
  geom_line(data = cell_towers_densities,
               mapping = aes(x = Elevation,
                             y = Frequency),
               color = "red")

ggplotly(g)# Create the HTML widget
```

This plot is revealing. The landscape data (represented by the black line and gray confidence interval) is left skewed and has two modes at c. 550 and 750 masl. In contrast, the cell tower data has a single dominant mode at c. 780 masl, and is slightly right skewed. From this visual investigation alone, we can see that the cell tower locations were unlikely to have been randomly sampled from the landscape as a whole.

### Mann-Whitney U test comparing non-normal distributions
Because neither of these distributions are statistically normal, we will use the nonparametric [Mann-Whitney U test](https://en.wikipedia.org/wiki/Mann%E2%80%93Whitney_U_test) (also known as a Wilcoxon test) to test whether the cell tower locations were likely a random sample of locations drawn from our study area. Again, we'll use Monte Carlo resampling to generate confidence intervals for our test statistic. Finally, we will output the test data to a comma-seperated values (CSV) file for inclusion in external reports.
```{r}

# Draw 999 random samples from the NED, and compute two-sample Wilcoxon tests (Mann-Whitney U tests)
my_county_Cell_MWU <- foreach(n = 1:99, .combine = rbind) %do% {
  my_county_sample <- my_county_NED_values %>%
    sample(nrow(cell_towers),
           replace = FALSE) %>%
    wilcox.test(x = cell_towers$`Elevation (m)`,
                y = .,
                alternative = "greater",
                exact = FALSE) %>%
    tidy() %>%
    tibble::as_tibble()
  
} %>%
  dplyr::select(statistic, p.value)

# Get the median test statistic and 95% confidence interval
my_county_Cell_MWU <- foreach::foreach(prob = c(0.025,0.5,0.975), .combine = rbind) %do% {
  my_county_Cell_MWU %>%
      dplyr::summarise_all(quantile, probs = prob)
} %>%
  t() %>%
  magrittr::set_colnames(c("Lower CI","Median","Upper CI")) %>%
  magrittr::set_rownames(c("U statistic","p-value"))

# Write output table as a CSV
my_county_Cell_MWU %T>%
  write.csv("OUTPUT/Mann_Whitney_results.csv")

```

The results of the Mann-Whitney U two-sample tests show that it is highly unlikely that the cell towers in Whitman county were randomly placed across the landscape (median U statistic = `r my_county_Cell_MWU["U statistic","Median"]`, median p-value = `r my_county_Cell_MWU["p-value","Median"]`). We can infer from the graphical analysis above that the cell towers were placed on unusually high places across the landscape.

## Conclusions
*R* is extremely powerful as a geo-analytic tool, and encountering sophisticated code for the first time can be daunting. But *R* can also be useful for basic exploratory spatial analysis, quick-and-dirty statistical testing, and interactive data presentation. I hope this short tutorial gave you a taste of how useful *R* can be.

## References cited